---
title: "Monopoly"
author: "Keen Zarate"
output:
  html_document: default
  pdf_document: default
---
```{r}
Monopoly<- read.csv("~/Downloads/MonopolyData.csv")
library(knitr)
library(kableExtra)
options(knitr.table.format = "html") 
```

My goal for this project is to simulate the game of *Monopoly* by 
implementing the steps for the paper *Take a Walk on the Boardwalk* 
by Abbott and Richey. I will be exploring this by verifying the four tables 
they used on their paper using markov chains. Note that the spaces used on 
the paper is less than what we will be using on this project, so data may 
not be 100% similar. Regardless, it would still work. 

Step 1: For this step, I will be making four transition matrices, roll, 
the go to jail, community chest, and the chance.
```{r}
cycleVec <- function(vec, k){
front <- tail(vec, k)
back <- head(vec, length(vec) - k)
c(front, back)
}

##Roll Trans
rolltrans <- matrix(0, nrow=40, ncol=40)
probs <- c(0,1/36,2/36,3/36,4/36,5/36,6/36,5/36,4/36,3/36,2/36,1/36,rep(0,28))
for(i in 1:40){
  rolltrans[,i] <- cycleVec(probs, i)
  }
colSums(rolltrans)

##Jail Trans
Jail <- matrix(0,nrow=40,ncol=40)
probsJ <- c()
for(i in 1:40){
  Jail[i,i] <- 1
}
Jail[31,31] <- 0
Jail[11,31] <- 1
colSums(Jail)

JR=Jail%*%rolltrans
```


## create the matrix for out Chance probabilities below.
```{r}
##Chance 
Chance <- matrix(0,nrow=40,ncol=40)
for(i in 1:40){
  Chance[i,i] <- 1
}
Chance[8,8] <- 6/16
Chance[1,8] <- 1/16
Chance[11,8] <- 1/16
Chance[25,8] <- 1/16
Chance[12,8] <- 1/16
Chance[40,8] <- 1/16
Chance[5,8] <- 1/16
Chance[13,8] <- 1/16
Chance[16,8] <- 2/16
Chance[6,8] <- 1/16

Chance[23,23] <- 6/16
Chance[1,23] <- 1/16
Chance[11,23] <- 1/16
Chance[25,23] <- 1/16
Chance[12,23] <- 1/16
Chance[40,23] <- 1/16
Chance[5,23] <- 1/16
Chance[13,23] <- 1/16
Chance[16,23] <- 2/16
Chance[6,23] <- 1/16

Chance[37,37] <- 6/16
Chance[1,37] <- 1/16
Chance[11,37] <- 1/16
Chance[25,37] <- 1/16
Chance[12,37] <- 1/16
Chance[40,37] <- 1/16
Chance[5,37] <- 1/16
Chance[13,37] <- 1/16
Chance[16,37] <- 2/16
Chance[6,37] <- 1/16

colSums(Chance)
```


Create the Community Chest probability matrix.
```{r}
CommunityChest <- matrix(0,nrow=40,ncol=40)
probsCC<- c()
for(i in 1:40){
  CommunityChest[i,i] <- 1
}
CommunityChest[3,3] <- 14/16
CommunityChest[1,3] <- 1/16
CommunityChest[11,3] <- 1/16

CommunityChest[28,28] <- 14/16
CommunityChest[1,28] <- 1/16
CommunityChest[11,28] <- 1/16

CommunityChest[34,34] <- 14/16
CommunityChest[1,34] <- 1/16
CommunityChest[11,34] <- 1/16

colSums(CommunityChest)

```

Table 1: 
```{r}
Table1 <- CommunityChest%*%Chance%*%Jail%*%rolltrans
eigenStuff1 <- eigen(Table1)
vecs1 <- eigenStuff1$vectors
ssVec1 <- vecs1[,1]
ssVec1 <- ssVec1/sum(ssVec1)
First <- sort(ssVec1, decreasing=TRUE, index.return=TRUE)
First1 <- matrix(0,nrow=40,ncol=3)
colnames(First1) <- c("Space","State","Frequency (%)" )
First1[,1] <- First$ix
States<-c(rep(0,40))
for(i in 1:40){
  States[i]<-as.character(Monopoly$Name[First$ix[i]])
}
First1[,2]<-States
First1[1,2]<-"Jail"
First1[,3]<-round(Re(First$x)*100,3)
Table1A<-kable(First1, "html") %>%
  kable_styling(bootstrap_options = "striped",full_width = F) %>%
  scroll_box(width = "100%", height = "400px")
Table1A
```
Conclusion: We can see that frequency values were pretty close to the ones in the paper. 
The paper used different space number for each state, so the frequency approximation may 
not exactly match the one Abbot and Richey found.  
    
## Step 2: Stay in jail

Next, after the game has played for a while, the player would want to stay in jail for as l
ong as they want. Only after three turns they have to move. 
I will be doing the same thing I did the previous step here, 
applying all the transition matrices that we already did in the previous steps.

```{r}
Chance <- matrix(0,nrow=42,ncol=42)
for(i in 1:42){
  Chance[i,i] <- 1
}
Chance[8,8] <- 6/16
Chance[1,8] <- 1/16
Chance[11,8] <- 1/16
Chance[25,8] <- 1/16
Chance[12,8] <- 1/16
Chance[40,8] <- 1/16
Chance[5,8] <- 1/16
Chance[13,8] <- 1/16
Chance[16,8] <- 2/16
Chance[6,8] <- 1/16

Chance[23,23] <- 6/16
Chance[1,23] <- 1/16
Chance[11,23] <- 1/16
Chance[25,23] <- 1/16
Chance[12,23] <- 1/16
Chance[40,23] <- 1/16
Chance[5,23] <- 1/16
Chance[13,23] <- 1/16
Chance[16,23] <- 2/16
Chance[6,23] <- 1/16

Chance[37,37] <- 6/16
Chance[1,37] <- 1/16
Chance[11,37] <- 1/16
Chance[25,37] <- 1/16
Chance[12,37] <- 1/16
Chance[40,37] <- 1/16
Chance[5,37] <- 1/16
Chance[13,37] <- 1/16
Chance[16,37] <- 2/16
Chance[6,37] <- 1/16

colSums(Chance)

CommunityChest <- matrix(0,nrow=42,ncol=42)
probsCC<- c()
for(i in 1:42){
  CommunityChest[i,i] <- 1
}
CommunityChest[3,3] <- 14/16
CommunityChest[1,3] <- 1/16
CommunityChest[11,3] <- 1/16

CommunityChest[28,28] <- 14/16
CommunityChest[1,28] <- 1/16
CommunityChest[11,28] <- 1/16

CommunityChest[34,34] <- 14/16
CommunityChest[1,34] <- 1/16
CommunityChest[11,34] <- 1/16

colSums(CommunityChest)

options(max.print=1000000)
rolltrans <- matrix(0, nrow=40, ncol=40)
probs <- c(0,1/36,2/36,3/36,4/36,5/36,6/36,5/36,4/36,3/36,2/36,1/36,rep(0,28))
for(i in 1:40){
  rolltrans[,i] <- cycleVec(probs, i)
  }

P2 <- matrix(0,nrow=42,ncol=42)
P2[1:40,1:40] <- rolltrans
rolltrans <- P2

rolltrans[42,41]<-5/6
rolltrans[11,42] <- 5/6
for(i in 1:6){
  rolltrans[11+2*i,41] <- 1/36
}
for(i in 1:6){
  rolltrans[11+2*i,42]<-1/36
}
colSums(rolltrans)

Jail <- matrix(0,nrow=42,ncol=42)
for(i in 1:42){
  Jail[i,i] <- 1
}
Jail[31,31] <- 0
Jail[41,31] <- 1

colSums(Jail)
```

Table 2
```{r}
Table2<-CommunityChest%*%Chance%*%Jail%*%rolltrans
eigenStuff2 <- eigen(Table2)
vecs2 <- eigenStuff2$vectors
ssVec2 <- vecs2[,1]
ssVec2 <- ssVec2/sum(ssVec2)
ssVecDif2<-ssVec2
ssVec2[11]<-ssVec2[41]+ssVec2[42]+ssVec2[11]
ssVec2<-ssVec2[1:40]


Second <- sort(ssVec2, decreasing=TRUE, index.return=TRUE)
Second1 <- matrix(0,nrow=40,ncol=3)
colnames(Second1) <- c("Space","State","Frequency (%)" )
Second1[,1] <- Second$ix
States2<-c(rep(0,40))
for(i in 1:40){
  States2[i]<-as.character(Monopoly$Name[Second$ix[i]])
}
Second1[,2]<-States
Second1[1,2]<-"Jail"
Second1[1,1] <- "11,41,42"
Second1[,3]<-round(Re(Second$x)*100,3)
Table2A<-kable(Second1, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  scroll_box(width = "100%", height = "400px")
Table2A
```

Step 3: Next we will use the rule that after rolling three doubles consecutively, 
the player is sent to jail. I will be using all the matrices already define above,
 with a bit of change. We have a 1/216 chance that the player will be send directly to jail.
```{r}
##Roll Trans
rolltrans <- matrix(0, nrow=40, ncol=40)
probs <- c(0,1/36,2/36,3/36,4/36,5/36,6/36,5/36,4/36,3/36,2/36,1/36,rep(0,28))
for(i in 1:40){
  rolltrans[,i] <- cycleVec(probs, i)
  }
colSums(rolltrans)

double<-matrix(0,42,42)
double[1:40,1:40]<-rolltrans
rolltrans<-double*215/216
for(i in 1:40){
  rolltrans[41,i]<-rolltrans[41,i]+1/216
}
rolltrans[42,41]<-5/6
rolltrans[11,42] <- 5/6
for(i in 1:6){
  rolltrans[11+2*i,41] <- 1/36
}
for(i in 1:6){
  rolltrans[11+2*i,42]<-1/36
}

stayinjail <- matrix(0,nrow=42,ncol=42)
probssty <- c()
for(i in 1:42){
  stayinjail[i,i] <- 1
}
stayinjail[31,31] <- 0
stayinjail[41,31] <- 1
```
Table 3:
```{r}
Table3 <- CommunityChest%*%Chance%*%stayinjail%*%rolltrans
eigenStuff3 <- eigen(Table3)
vecs3 <- eigenStuff3$vectors
ssVec3 <- vecs3[,1]
ssVec3 <- ssVec3/sum(ssVec3)
ssVecDif3<-ssVec3
ssVec3[11] <- ssVec3[42]+ssVec3[41]+ssVec3[11]
ssVec3<-ssVec3[1:40]

Third <- sort(ssVec3, decreasing=TRUE, index.return=TRUE)
Third2 <- matrix(0,nrow=40,ncol=3)
colnames(Third2) <- c("Space","State","Frequency (%)" )
Third2[,1] <- Third$ix
States3<-c(rep(0,40))
for(i in 1:40){
  States3[i]<-as.character(Monopoly$Name[Third$ix[i]])
}
Third2[,2]<-States3
Third2[1,2]<-"Jail"
Third2[1,1] <- "11,41,42"
Third2[,3]<-round(Re(Third$x)*100,3)
Table3A<-kable(Third2, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  scroll_box(width = "100%", height = "1000px")
Table3A
```


## Step 4: Game analysis
```{r}
Table4<-matrix(0,nrow=8,ncol=4)
colnames(Table4)<-c("Group color","Group frequency(%)","Group Value per roll ($)","Rolls to break even")
Table4[,1]<-c("Purple (Baltimore,Mediterranean)","Light Blue (Oriental,Vermont,Connecticut)","Maroon (St. Charles,States,Virginia)","Orange (St. Jaames, Tennesse,  NY)","Red (Kentucky,Indiana, Illinois)","Yellow (Atlantic, Ventnor, Marvin Gardens)","Green (Pacific,N. Carolina, Pennsylvania)","Dark Blue (Park Place, Boardwalk)")
```

###Steady-state Probabilities
* What are the steady state probabilities for each color group? 
(just add up the probabilies for all the properties in the group).
```{r}
Table4[,2]<-c(rep(0,8))
for(i in 1:40){
  for(j in 1:8){
    if(Monopoly$Group[i]==j){
      Table4[j,2]<-as.numeric(Table4[j,2])+Re(round((ssVec3[i])*100,3))
    }
  }
}
```



###Earn per roll
* Assuming the group is fully developed, how much does it earn per roll? 
(Multiply the total rent times the probability).
```{r}
Table4[,3]<-c(rep(0,8))
for(i in 1:40){
  for(j in 1:8){
    if(Monopoly$Group[i]==j){
      Table4[j,3]<-as.numeric(Table4[j,3])+Re(round((ssVec3[i]*Monopoly$Rent[i]),3))
    }
  }
}
```


##Rolls to Break Even
* If you know how much a property earns per roll, 
how many rolls does it take to "break even" for the property? 
(Divide total development cost by earnings per roll).
```{r}
Property<-c(rep(0,8))
for(i in 1:40){
  for(j in 1:8){
    if(Monopoly$Group[i]==j){
      Property[j]<-Property[j]+Monopoly$Cost[i]+Monopoly$Total[i]
    }
  }
}
Table4[,4]<-round(Property/as.numeric(Table4[,3]),1)
```

Table 4: Game Analysis of *Color Groups*
```{r}
Table4A<-kable(Table4, "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  scroll_box(width = "100%", height = "400px")
Table4A
```
Again above, we confirm what we were shown from the paper.


Conclusion: My implementation gave values that were close to the values found from the paper. 
I think the first thing I would change on this code is my implentation of the Chance and the 
Community Chest, which I think could be more efficient. Another thing we could change or add 
on the exploration of this game is possibly including the fact that instead of just staying in jail, 
a player may want to leave jail as soon as possible by rolling doubles, pay a fine, or getting the Jail free card.  