---
title: "Traveling Salesperson Problem"
author: "Keen Zarate"
---
###Goal:
The goal for this project is to use *Simulated Annealing* to figure out the shortest possible 
route that a traveling salesperson could take when visiting multiple cities and returning to 
the initial city. In addition, we will see an how the simulated annealing can produced routes 
that takes into account cost of travel. In this case, if it is more expensive to travel 
north-south than it is east-west. We'll see how it works!!!

###Implementation Setup
Below is just creating N random points which represents the "cities" the salesman would travel. 
```{r}
N <- 55
xyvals <- matrix(runif(2*N, -10, 10), nrow=N)
library(ggplot2)
ggplot(data=NULL, aes(x=xyvals[,1], y=xyvals[,2])) +
 geom_point(size=2, color="blue")
```

###Tours, Permutations, and Distances
Below is to simulate just random routes the this salesman would follow. Probably wouldn't be the best one. 
```{r}
tour <- 1:N
ggplot(data=NULL, aes(x=xyvals[,1], y=xyvals[,2])) +
geom_point(size=2, color="blue") +
geom_polygon(color="purple", fill=NA)

```

###Length of Tour
Below is the function that would calculate the length of each tour by picking some of those random points. 
```{r}
distPair <- function(i, j, tour){
x1 <- xyvals[tour[i],1]
x2 <- xyvals[tour[j],1]
y1 <- xyvals[tour[i],2]
y2 <- xyvals[tour[j],2]
return(sqrt((x1-x2)^2+(y1-y2)^2))
}
```

Let's check?
```{r}
distPair(1,2,tour)
```

###Distance of entire route
Below is the function that will calculate the distance of the entire tour of the salesman.
```{r}
distTour <- function(tour){
tot <- 0
N <- length(tour)
for(i in 1:N){
  if (i==N){
    tot <- tot + distPair(N,1,tour)
  } else {
    tot <- tot + distPair(i,(i+1),tour)
  }
}
tot
}
```

Let's check:
```{r}
trivialTour <- 1:N
distTour(trivialTour)
```

###Random Tour
We can also simulate random tours
```{r}
randTour <- sample(1:N, N, rep=FALSE)
randTour
```

```{r}
distTour(randTour)
```

###Plot of *random tour*
```{r}
plotTour <- function(tour, ptSize=2){
ggplot(data=NULL, aes(x=xyvals[tour,1], y=xyvals[tour,2])) +
geom_point(size=ptSize, color="blue") +
geom_polygon(color="purple",fill=NA)+
    geom_text(aes(label=1:N),hjust=-1)
}
plotTour(randTour)
```

###Simulated Annealing and the TSP
So,below is a way to optimize those routes.. to get a better and more effiecient travel for the salesman. We will be using simulated annealing for this process and let's see what it will produce us.

###N Random Points
Below are just random poitns created, just like what we did from earlier.
```{r}
N <- 55
xyvals <- matrix(runif(2*N,-10,10),nrow=N)
library(ggplot2)
ggplot(data=NULL,aes(x=xyvals[,1],y=xyvals[,2]))+
  geom_point(size=2,color="blue")+
  geom_text(aes(label=1:N),hjust=-1)
```

###Trivial Tour
This is just a possibility of routes we could take, this is super messy though, 
we want something less crazier than this. 
```{r}
trivialTour <- 1:N
plotTour(trivialTour)
```

Compute the initial distance. This is the only time you need to use *disTour* function.
```{r}
dist0 <- distTour(trivialTour)
dist0
```

###Simulating Annealing
Finally! We will now use Simulated Annealing.
We begin by defining important parameters. M is the number of iterations, 
we need to define sigma and the the decreasing factor. 
```{r}
M <- 10000
sig2 <- 0.1
decFac <- 0.9999
```

Next: The important part of the puzzle. In this for loop below, 
we are expressing what where we want the tour to go and basically using the values 
from above to give an optimized values for those tour.
```{r}
for(m in 1:M){
  place <- sample(1:N,2,rep=F)
  pos0 <- min(place)
  pos1 <- max(place)
  if(pos0 == 1 & pos1 == N){
  next
}

  pos0old <- pos0 - 1
  pos1new <- pos1 + 1

  if(pos0 == 1){
  pos0old <- N
  }

  if(pos1 == N){
  pos1new <- 1
  }

  distLost <- distPair(pos0,pos0old,trivialTour) + distPair(pos1,pos1new,trivialTour)
  distGain <- distPair(pos0old,pos1,trivialTour) + distPair(pos0,pos1new,trivialTour)
  deltaDist <- distGain - distLost
  dist1 <- dist0 + deltaDist

  

  trivialTour[pos0:pos1]<- trivialTour[pos1:pos0]

  rho <- exp(-deltaDist/sig2)
  if(runif(1,0,1) < rho){
  dist0 <- dist1
  } else{
   trivialTour[pos0:pos1]<- trivialTour[pos1:pos0]
  }
  sig2 <- sig2*decFac
}

plotTour(trivialTour)
dist0

print(c(deltaDist,sig2))
```

Conclusion: As we can see, it's a lot neater than the previous random tours we just did. 
The routes are not 100% optimized. Certain points could possibly be flipped to get better 
routing, but again, we can possibly change the values of our sigma and also our decreasing factor to improve outcome. 

###Extension:
Suppose north-south travel is more expensive than east-west travel. Let's modify our distance function so
that the difference between y-values is multiplied by some constant c. This is now the function to minimize
along the tour.

Essentially for this extension, we just need to change one thing from the 
process that we already had earlier -- adding a cost value. 
Let's implement this using our codes from earlier.

```{r}
N <- 55
trivialTour <- 1:N
M <- 10000
xyvals <- matrix(runif(2*N,-10,10),nrow=N)
sig2 <- 0.1
decFac <- 0.9999
```

```{r}
plotTour <- function(tour, ptSize=2){
ggplot(data=NULL, aes(x=xyvals[tour,1], y=xyvals[tour,2])) +
geom_point(size=ptSize, color="blue") +
geom_polygon(color="purple",fill=NA)+
    geom_text(aes(label=1:N),hjust=-1)
}
```

###Cost Function
```{r}
c <- 15
costPair <- function(i, j, tour){
x1 <- xyvals[tour[i],1]
x2 <- xyvals[tour[j],1]
y1 <- xyvals[tour[i],2]
y2 <- xyvals[tour[j],2]
return(sqrt((x1-x2)^2+c*(y1-y2)^2))
}
```

###Cost of entire route
```{r}
costTour <- function(tour){
tot <- 0
N <- length(tour)
for(i in 1:N){
  if (i==N){
    tot <- tot + costPair(N,1,tour)
  } else {
    tot <- tot + costPair(i,(i+1),tour)
  }
}
tot
}
dist0 <- costTour(trivialTour)
```


###Simulated Annealing Part
Let's do this!!! See what happens...
```{r}
for(m in 1:M){
  place <- sample(1:N,2,rep=F)
  pos0 <- min(place)
  pos1 <- max(place)
  if(pos0 == 1 & pos1 == N){
  next
}

  pos0old <- pos0 - 1
  pos1new <- pos1 + 1

  if(pos0 == 1){
  pos0old <- N
  }

  if(pos1 == N){
  pos1new <- 1
  }

  costLost <- costPair(pos0,pos0old,trivialTour) + costPair(pos1,pos1new,trivialTour)
  costGain <- costPair(pos0old,pos1,trivialTour) + costPair(pos0,pos1new,trivialTour)
  deltaCost <- costGain - costLost
  dist1 <- dist0 + deltaCost

  

  trivialTour[pos0:pos1]<- trivialTour[pos1:pos0]

  rho <- exp(-deltaCost/sig2)
  if(runif(1,0,1) < rho){
  dist0 <- dist1
  } else{
   trivialTour[pos0:pos1]<- trivialTour[pos1:pos0]
  }
  sig2 <- sig2*decFac
}

plotTour(trivialTour)
dist0

print(c(deltaCost,sig2))
```

###Conclusion: 
It looks like the travel is more towards east-west which makes sense because north-south 
is more expensive. It looks like is doing better also on trying to optimize the best route 
possible just looking at the points in comparison to earlier where only the distance was studied. 
It looks so much neater and "cheaper" to travel this way. 
It seems to be doing better on optimizing the route than from earlier. Cool stuff!


###Limitation and Future Work
I think one limitation of this simulated annealing project is that, 
sigma and the decreasing factor needs to me manipulated a lot to get the best results. 
I think that there could be a way to not get a result until the graph shows that best routes. 
Like it would just keep iterating until it finds the best possible routes. 
Also, may be for future modeling, may be simulated annealing could be done better and instead 
of having sigma and some decreasing factor, like I suggested earlier could have some kind of 
code that allows for better optimization of routes and iterates multiple times until we get t
he better result. I think that overall, the simulated annealing does help on this. 
I think one thing would be interesting to look at for this TSP could be like patterns of travels around 
the world, comparing cold countries vs warm and trying to see where likely people would travel at a 
given year if climate/weather were taken into account. I read a paper that does this which studied 
temperature and travel using simulated annealing. 